Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-07-07 analysis/paper/paper.qmd

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A survey of reproducibility reviews for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

\newpage

>>>>

<<<<<<< HEAD
---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A survey of reproducibility reviews for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date. To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025. This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator. Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.

# compute diversity of all citations for each article (habitat)
shannon_per_item <- 
  ref_list7 %>% 
  group_by(id, x) %>% 
  tally() %>% 
  mutate(n_in_article = n) %>% 
  select(-n) %>% 
  left_join(all_cited_items) %>% 
  mutate(p_i = n / sum(n, na.rm = TRUE)) %>% 
  mutate(p_i_ln = log(p_i)) %>% 
  group_by(id) %>% 
  summarise(shannon = -sum(p_i * p_i_ln, na.rm = TRUE)) %>% 
  mutate(id = as.numeric(id)) %>% 
  arrange(id)  %>% 
  left_join(items_df)

The PCA results in @fig-pca show that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is a reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. Variation in PC2 is influenced by the number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeological Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.

Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results. The most common issue was an incomplete compendium. This ranges from missing data files down to missing lines of code. In most cases this can be attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions. Unspecified or under-specified dependencies is another common issue that prevents code from running. This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations (e.g. dplyr for R or numpy for Python). If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change the way they behave as the developers update their package. Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.

\newpage
=======
---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A survey of reproducibility reviews for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

\newpage
>>>>>>> e6713de47690905ff4d77466e22765f2b1ce537e
